\section{Conclusion and Outlook}

In this work, we analyzed \sort for sorting aligned DNA-Read files, specifically BAM files. We found that the most time-consuming part of sorting is compression and writing of output and temporary files. To reduce the runtime of \sort, we proposed setting a higher limit for the number of temporary files concurrently stored on disk, analyzed alternative implementations of the compression library used by SAMtools, and examined the impact of IO requirements on the runtime of \sort.

Setting a higher limit for temporary files concurrently stored on disk reduces the number of merges \sort performs, leading to lower runtimes when sorting large files with limited memory.
By using libdeflate as the compression library, which is automatically the case if SAMtools is installed via Bioconda, we achieved a single-thread speedup of 2.3 compared to using zlib. On 16 threads, this results in a speedup of 1.6.
By utilizing Unix pipelines, we can remove the output compression of \sort, achieving a speedup of 1.8 when \sort is piped to SAMtools \texttt{view} (SAMtools \texttt{view} with zlib compression at level 6).
To increase user awareness of better compression options, we recommended implementing warnings if zlib is used instead of libdeflate, and if the output of \sort is piped but still compressed. 

Combining our optimizations and using libdeflate for decompression, igzip with compression level 1 for compression of temporary and output files, and an increased limit for temporary files, we could archive a speedup of 6 compared to \sort with the default zlib compression on compression level 6 for single-threaded sorting of a 215\,GiB BAM file utilizing 32\,GiB of memory. Compared to \sort with libdeflate compression on compression level 6, this equals a speedup of 3.\footnote{
Runtimes single-threaded: with zlib (6): 41395\,s, with libdeflate (6): 20815\,s, with libdeflate decompression and igzip (1) compression: 6887\,s. 
Runtimes utilizing 16 additional threads: with zlib (6): 4287\,s, with libdeflate (6): 3488\,s, with libdeflate decompression and igzip (1) compression: 2217\,s. (Compression levels in parentheses)
}
Utilizing 16 additional threads for sorting a 215\,GiB BAM file making use of 32\,GiB of memory, our optimizations lead to a speedup of 2 compared to \sort with the default zlib compression on compression level 6, and a speedup of 1.5 compared to \sort with libdeflate compression on compression level 6. 

These improvements for the runtime of \sort represent an important contribution to the field of bioinformatics, considering the widespread adoption of \sort evidenced by its over 5,000 citations and over 5.1 million downloads through Bioconda.

Future projects can investigate the merging process further, as this appears to be a bottleneck for very fast compression libraries. Additionally, they can implement igzip support into SAMtools and its file operation library HTSlib, as igzip has lower runtimes than libdeflate. Furthermore, the merging strategy of \sort can be enhanced by writing temporary files not only to half the limit for temporary files but to the limit minus existing "big files", which are results of previous merges, thereby halving the merges for the first few merges.