\section{Input/Output} 
Input and Output can also be constraints of the sorting process. As the internal mechanisms of SAMtools usually work very fast and are highly parallel, but have to process huge amounts of data, input and output devices can also limit the computation speed.

\subsection{IO}
In contrast to the used compression library, often the IO devices can hardly be changed. However, in some cases there are possibilities to speed up the process.

\subsubsection{Pipelining} enables all advantages of working with streams. For SAMtools \texttt{sort}, this has the following consequences: Records can be read as they are produced by the previous command. This can be done without compression and without having to write files to the disk using UNIX pipelines. Note that the process generating the input for SAMtools \texttt{sort} is most likely halted once the memory limit of \texttt{sort} is reached. This is due to the buffer offered by the pipeline being full and only emptied again, after a temporary files is written. Looking at the output of sort, there is also no need for compression and writing to disk if piped. 

\subsubsection{Prefixes for temporary files} can be set via the "\texttt{-T}" parameter. If no prefix is specified, temporary files are written into the same directory as the output. If the output is to standard output, they are placed in the current working directory. Here, a directory on a fast disk should be chosen. As temporary files are deleted automatically after successful sorting, only at the time of sorting the disks capacities are used. However, it is important to remember, that temporary files are less compressed than regular input and output files. In combination, they require about 20\% more disk space, than the input file. Moreover, if the hard drive is very fast and additionally offers enough storage space, the compression of the intermediate files can be omitted. Unfortunately, this is not possible without changing the source code at the moment. This can be done by simply replacing the parameter \texttt{mode} of the first call of \texttt{bam\_merge\_simple} in the \texttt{bam\_sort\_core\_ext} method which is located in \texttt{bam\_sort.c}. Current values are, depending on the existence of a position too large to be stored in a BAM file, "\texttt{wzx1}" for BGZF compressed SAM files on compression level 1 and "\texttt{wbx1}" for BAM files with compression level 1. Those can be changed to "\texttt{w}" for SAM files and "\texttt{wbx0}" for uncompressed BAM files.