\section{Related Work}

\subsection{Sorting Tools}
SAMtools is the most commonly used tool for manipulating SAM and BAM files, which store information on aligned DNA-Reads. With over 5.1~Million downloads (05.2014), it is the most downloaded package from the Bioconda~\cite{the_bioconda_team_bioconda_2018} channel, which is a common source for bioinformatic tools in the package and dependency manager Conda. However, there are some alternatives for sorting aligned DNA-Read files, such as sambamba~\cite{tarasov_sambamba_2015}, Picard~\cite{Picard2019toolkit} and NovoSort~\cite{noauthor_novosort_nodate}. Picard, written in Java, provides a simpler interface for sorting SAM and BAM files, but its functions are limited compared to sort, and it does not prioritize efficiency. NovoSort is a commercial program and, therefore, I could not test it. Sambamba, written in the D programming language~\cite{alexandrescu_d_2010}, aims to provide a subset of SAMtools functionality, including sort, but with greater efficiency through higher parallelization. Sambamba reached this goal during its implementation in 2012 as SAMtools. However, with SAMtools introducing parallelism in version 1.4 in 2017, this advantage diminished.\\

\subsection{Compression Methods and Ouput Files}
Compression using BGZF, which is part of the specification of BAM files utilizes GZIP. As a general purpose compression method, GZIP is not specialized in compression DNA data. Hence, a variety of compression algorithms have been developed specifically for DNA data~\cite{hosseini_survey_2016}. These algorithms aim to minimize the resulting file size, compression speed, memory usage during compression or decompression, or decompression speed, but typically focus primarily on minimizing the resulting file size. Compression algorithms are divided in \textit{reference-based} and \textit{reference-free} methods. \\

Reference-based algorithms utilize the reference sequence a DNA-Read is aligned too. By only storing differences between each DNA-Read sequence and the reference sequence, they archive better compression. NGC~\cite{popitsch_ngc_2013} is an example for this, which also splits the kind of data, namely the sequences, sequence names, read qualities and so on in different blocks and compresses them seperately. Other algorithms like Goby~\cite{campagne_compression_2013}, DeeZ~\cite{hach_deez_2014}, and CRAM~\cite{fritz_efficient_2011} are also reference-based, but keep the advantage of enabling random access through index files. CRAM (Compressed Reference-oriented Alignment Map) is also built into HTSlib, the library SAMtools utilizes for file operations, and the user can choose to use it as output for \sort. NGC, Goby and CRAM also offer lossy compression. Lossy compression mainly removes meta information or applies lossy compression to the read qualities of DNA-reads, as their exact value is not always needed, it does not depend on the alignment and is thus hard to compress lossless. Although reference compression methods for SAM files are still actively developed~\cite{banerjee_abridge_2022}, they lack interoperability, as the compression methods are not widely used. An exception to this is CRAM, which is supported by HTSlib. Although CRAM files offer several advantages over BAM files, BAM files remain more popular due to their wider support among software tools.\\

Reference-free compression methods like BGZF, which is used in BAM files per specification, offer the advantage of not needing the reference sequence for compression and decompression. In addition, reference-based compression methods often require the DNA-Reads to be sorted. As this is not necessary for reference-free compression methods, they are more flexible. However, the only commonly used reference-free compression method is the BAM format, which uses binary encoding for the bases a DNA sequence consists of and BGZF compression. \\

In summary, the BAM file format employing BGZF compression, which internally utilizes GZIP, is the most commonly used format for storing aligned DNA-Reads. Given its widespread use, optimizing the speed of generating BAM files using \sort is essential. However, the increasing adoption of the CRAM format might lessen the future significance of BAM file generation.

